Choose a License

### 2025-01-22 choose_license.qmd
Simplify after consultation

A license is a legal document that regulates what _others_ are allowed to do with a copyright-protected work -- the licenses we discuss are non-exclusive licenses and do not limit the author or copyright holder in their rights. And while you _could_ write your own license, there are already many pre-formulated licenses available to choose from and to apply to your work.^[In fact, you should not create your own license if the purpose is to share your work freely and openly with others. By using a boilerplate license others know what to expect, also because they have been [tested](https://sfconservancy.org/copyleft-compliance/) in [courts](https://legaldb.creativecommons.org/).]

@@

::: {#wrn-exclusive-rights .callout-warning}
### Giving Publishers Exclusive Rights

Some journals offer to publish your article under a Creative Commons license, but still demand an exclusive publishing and distribution license or a copyright assignment from you. This would give them more rights than the readers of the article have through the respective Creative Commons license [@Rumsey2022] and exceeds by far what is necessary to make publication possible [@Suber2022]. Consequently, authors should oppose this practice and grant publishers the same rights that every other reader of the article has. To facilitate self-archiving, one can also modify the contract with publishers via a rights retention statement [@UKRN2023] / author's addendum [@SPARC2006]. If you have published a closed-access paper before, you can consult [ShareYourPaper](https://shareyourpaper.org/) for legal options to still make it available free of charge to readers.

If your chosen publisher insists on an exclusive license, you may at least retain the copyright for your figures -- follow the guide "Retaining copyright for figures in academic publications to allow easy citation and reuse" by @Elson2016 to learn how to do that.
:::

@@

Note, that for data there is, at least in principle, the possibility that a different license applies to the individual entries than to the collective database. For example, if you were to create a database of artworks by others, those artworks would be licensed individually as chosen by the artists, but the license for the database as a whole could be chosen by you. The latter includes the selection of entries, field names and any _sui generis_ database rights. However, if the content was created by you, we recommend you to choose the same license for both content and database. Factual data (like measurements or metadata) should be licensed under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) [@Villa2016Copyleft; @Villa2016CC0] -- otherwise, consult @nte-other-work-types for some caveats.

::: {#fig-flowchart-simple}
```{mermaid}
flowchart TB
  start("We want<br>to choose<br>a license.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>concerning license?")

@@

One should take special care when applying license to the following types of works:

- __fonts:__ Copyleft licenses applied to fonts can be a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. The [SIL Open Font License\ 1.1](https://openfontlicense.org/) is the _de facto_ standard license for fonts, requiring that derived fonts, if published, have to be put under the same license. It doesn't require attribution for usage (e.g., when embedded in a document), but forbids that the font is sold by itself. Selling the font is allowed, however, if it is bundled with other software.
- __templates and TeX packages:__ If a template or, for this purpose, a $\TeX$ package, is licensed under a copyleft license such as the (A)GPLv3, every work that is a derivative of the template or that uses the $\TeX$ package has to put under the same license, if shared [@Koppor2016]. And if a document contained source code covered by the (A)GPLv3, the same license would also apply to the document [@MadHatter2022].
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access (its Anti-DRM provision) if shared [see @Poole2017]. Similarly, following an example from @Matt2009, if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to be licensed under [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or another license).
:::

@@

- __Attribution__ means whether recipients of your work are required to provide attribution to you. For software licenses, the attribution is part of a copyright notice. Note that even if attribution is not a requirement of the license, good scientific practice demands that appropriate citations are made.
- __State changes__ means that recipients need to indicate if changes were made.
- __Anti-DRM__ means that when others share your work they are not allowed to apply technological measures restricting anything that the license permits. Implementing such restrictive measures is also referred to as **d**igital **r**ights **m**anagement (DRM).

@@

As indicated before, a free/open license must allow creating derivative works and must allow exercising the rights granted by it _for any purpose_, including commercial use. There are a few good resources on the reasoning behind that [e.g., see @Klimpel2013; @ODI2015; @Stallman2022NoLimit; @Moller2023; @Vezina2020; @WikimediaCommons2009], but we would like to highlight one reason in particular.

@@

However, applying a restriction such as only permitting non-commercial use or not allowing derivative works creates _silos_ of works which are mutually incompatible with each other. Put differently, one cannot share a remix of two works where one is licensed under CC\ BY-SA\ 4.0 and the other only allows non-commercial use. In order to avoid silos, one should only choose licenses which are compatible with the dominant copyleft license in the respective community [@Lammerhirt2017; @Wheeler2014]. If you would like to learn more about the different types of compatibility, we recommend you to read the article "A Quick Guide to Software Licensing for the Scientist-Programmer" by @Morin2012. The following diagram provides an overview of the compatibility of various licenses:

!["Open data-capable license interoperability" by Robbie Morrison [-@Morrison2024] licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/)](images/Morrison2024.jpg){.lightbox}
:::

@@

Having selected the licenses of your choice -- again, you might need multiple ones depending on the types of works your project contains --, we encourage you to read through the full license text (or at least a legal summary) to understand their effect. Then, you can record the license of existing content and apply the licenses of your own contributions. Mostly, this just means indicating which file or folder is covered by which license, usually in the project's README (among other places), whose creation will be discussed [later](make_readme.qmd). Notably, no sort of registration is required.

@@

You may have noticed that we mostly refer to licenses using a name _and_ a version number. This is because the organizations that created the licenses sometimes publish updated versions to accommodate for developments in copyright law and the communities that use the licenses. For example, the Creative Commons licenses (that start with `CC`) were first published in 2002. Since then, the possibility to relicense under later ([v2.0](https://creativecommons.org/weblog/entry/4216#:~:text=Share%20Alike%20Across%20Borders), 2004) and compatible licenses ([v3.0](https://creativecommons.org/weblog/entry/7249), 2007) has been added, a 30-day window to correct license violations has been established to combat [copyleft trolls](https://commons.wikimedia.org/wiki/Commons:Copyleft_trolling), and _sui generis_ database rights are covered explicitly ([v4.0](https://creativecommons.org/weblog/entry/40768), 2013). There are many more [subtle differences between license versions](https://wiki.creativecommons.org/wiki/License_Versions), therefore it is important to indicate which license version exactly one is referring to, as the license of a work does not "update" automatically. CC0, officially introduced in 2009, is still at version [1.0](http://www.creativecommons.org/weblog/entry/13304).

@@

```{.bash .code-overflow-wrap filename="Terminal"}
reuse annotate --copyright="Kristen Gorman" --license="CC0-1.0" data.csv
```

You will notice that this creates another file called `data.csv.license` containing the relevant information:

```{.yml filename="data.csv.license"}
SPDX-FileCopyrightText: 2024 Kristen Gorman

@@

  subgraph content["<strong>License for content</strong>"]
  existing_license_content --"No, we created the content<br>entirely by ourselves."--> facts("Entries are facts<br>(like measurements<br>or metadata)?")
  existing_license_content --"Yes, it was<br>shared under a<br>free/open license."--> use_existing_license_content["<em>Use that license</em>"]
  facts --"Yes"--> cc0_content_metadata["CC0 1.0"]
  facts --"No"--> choose_license["<em>Consult flowchart for<br>software, writing,<br>image, audio, and video</em>"]
  end

  subgraph database["<strong>License for database</strong>"]
  choose_license --> switch_license["<em>Depending on<br>content license</em>"]
  use_existing_license_content --> switch_license

>>>>

::: {#nte-situation-lmu .callout-note collapse="true"}
### For researchers at LMU

If you are an employee at LMU and your funder does not mandate the use of a particular license, the decision of which license to use lies with the holder of your chair (_Lehrstuhlinhaber\*in_). The only exception is the case that you developed an invention you would like to exploit commercially with a patent. Then, you are required to officially report your invention to the IP management team of the LMU, which will work out the next steps with you.
:::

A license is a legal document that regulates what _others_ are allowed to do with a copyright-protected work -- the licenses we discuss do not limit the author or copyright holder in their rights. And while you _could_ write your own, there are already many pre-formulated licenses available to choose from and to apply to your work.^[In fact, you should not create your own license if the purpose is to share your work freely and openly with others. By using a boilerplate license others know what to expect, also because they have been [tested](https://sfconservancy.org/copyleft-compliance/) in [courts](https://legaldb.creativecommons.org/).]

@@

::: {#cau-exclusive-rights .callout-caution}
### Giving Publishers Exclusive Rights

When you have your article published under a Creative Commons license, some publishers still demand an exclusive publishing and distribution license or a copyright assignment. This would give them more rights than the readers of the article have through the chosen Creative Commons license [@Rumsey2022] and exceeds by far what is necessary to make publication possible [@Suber2022]. Consequently, authors should oppose this practice and grant publishers the same rights that every other reader of the article has. If your chosen publisher insists on an exclusive license, you may at least retain the copyright for your figures -- follow the guide "Retaining copyright for figures in academic publications to allow easy citation and reuse" by @Elson2016 to learn how to do that.

If you have published a closed-access paper before, you can consult [ShareYourPaper](https://shareyourpaper.org/) for legal options to still make it available free of charge to readers.
:::

@@

Note, that for data there is, at least in principle, the possibility that a different license applies to the individual entries than to the collective database. For example, if you were to create a database of artworks by others, those artworks would be licensed individually as chosen by the artists, but the license for the database as a whole could be chosen by you. The latter includes the selection of entries, field names and any _sui generis_ database rights. However, if the content was created by you, we recommend you to choose the same license for both content and database. Factual data (like measurements or metadata) should be licensed under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) -- otherwise, consult @nte-other-work-types for some caveats.

::: {#fig-flowchart-simple}
```{mermaid}
flowchart TB
  start("We want<br>to choose<br>a license.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>regarding license?")

@@

One should be cautious about the restrictions of licenses applied to the following types of works:

- __fonts:__ Copyleft licenses applied to fonts can be a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. If the intent is that only derived fonts, if published, have to be put under the same license, the [SIL Open Font License\ 1.1](https://openfontlicense.org/) is an appropriate choice. Note, however, that it doesn't require attribution for usage. If no copyleft mechanism is intended, [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) also works for fonts.
- __templates and LaTeX packages:__ If a template or, for this purpose, a LaTeX package, is licensed under a copyleft software license such as the [AGPLv3](https://choosealicense.com/licenses/agpl-3.0/), every work that is a derivative of the template or that uses the LaTeX package has to put under the same license, if shared [@Koppor2016]. And if a document contained source code covered by the GPLv3, the same license would also apply to the document [@MadHatter2022].
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access (its Anti-DRM provision) if shared [see @Poole2017]. Similarly, following an example from @Matt2009, if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to licensed under CC\ BY-SA\ 4.0 (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or even another license).
:::

@@

- __Attribution__ means whether recipients of your work are required to provide attribution to you. For software licenses, this is called a copyright notice. Note that even if attribution is not a requirement of the license, good scientific practice demands that appropriate citations are made.
- __State changes__ means that recipients need to indicate if changes were made.
- __Anti-DRM__ means that when others share your work they are not allowed to apply technological measures restricting anything that the license permits. DRM is an abbreviation for **d**igital **r**ights **m**anagement.

@@

As indicated before, a free/open license must allow creating derivative works and must allow exercising the rights granted by it _for any purpose_, including commercial use. There are a few good resources on the reasoning behind that [e.g., see @Klimpel2013; @ODI2015; @Stallman2022NoLimit; @Moller2023], but we would like to highlight one reason in particular.

@@

However, applying a restriction such as only permitting non-commercial use or not allowing derivative works creates _silos_ of works which are mutually incompatible with each other. Put differently, one cannot share a remix of two works where one is licensed under CC\ BY-SA\ 4.0 and the other only allows non-commercial use. In order to avoid silos, one should only choose license which are compatible with the dominant copyleft license in the respective community [@Lammerhirt2017; @Wheeler2014]. If you would like to learn more about the different types of compatibility, we recommend you to read the article "A Quick Guide to Software Licensing for the Scientist-Programmer" by @Morin2012. The following diagram provides an overview of the compatibility of various licenses:

!["Open data-capable license interoperability" by Robbie Morrison licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/). Taken without modification from @Morrison2024](images/Morrison2024.jpg){.lightbox}
:::

@@

Having selected the licenses of your choice -- again, you might need multiple ones depending on the types of works your project contains --, we encourage you to read through the full license text (or at least a legal summary) to understand their effect. Then, you can record the license of existing content and apply the licenses of your own contributions. Mostly, this just means indicating which license applies to which file or folder, usually in the project's README (among other places), whose creation will be discussed [later](make_readme.qmd).

@@

You may have noticed that we mostly refer to licenses using a name _and_ a version number. This is because the organizations that created the licenses sometimes publish updated versions to accommodate for developments in copyright law and the communities that use the licenses. For example, the Creative Commons licenses (that start with `CC`) were first published in 2002. Since then, the possibility to relicense under compatible licenses has been added ([v3.0](https://creativecommons.org/2007/02/23/version-30-launched/)), a 30-day window to correct license violations has been established to combat [copyleft trolls](https://commons.wikimedia.org/wiki/Commons:Copyleft_trolling), and _sui generis_ database rights are covered explicitly ([v4.0](https://creativecommons.org/version4/)). There are many more [subtle differences between license versions](https://wiki.creativecommons.org/wiki/License_Versions), therefore it is important to indicate which license version exactly one is referring to, as the license of a work does not "update" automatically.

@@

```{.bash .code-overflow-wrap filename="Terminal"}
reuse annotate --copyright="Kristen Gorman" --license="CC0-1.0" Data.csv
```

You will notice that this creates another file called `Data.csv.license` containing the relevant information:

```{.yml filename="Data.csv.license"}
SPDX-FileCopyrightText: 2024 Kristen Gorman

@@

  subgraph "License for individual data entries (content)"
  existing_license_content --"No, we created the content<br>entirely by ourselves."--> facts("Entries are facts<br>(like measurements<br>or metadata)?")
  existing_license_content --"Yes, it was<br>shared under a<br>free/open license."--> use_existing_license_content["<em>Use that license</em>"]
  facts --"Yes"--> cc0_content_metadata["CC0 1.0"]
  facts --"No"--> choose_license["<em>Consult flowchart for<br>software, writing,<br>image, audio, and video</em>"]
  end

  subgraph "License for combination of data (database)"
  choose_license --> switch_license["<em>Depending on content license</em>"]
  use_existing_license_content --> switch_license
